---
title: "Taxonomy Bar Plot Tutorial"
author: "Avril Metcalfe-Roach"
date: "`r Sys.Date()`"
output: html_document
---

# Background

Taxonomic bar plots show the relative abundances of the most abundant taxa. Taxa
under 1% abundance are often grouped as otherwise they are not visible/there are
too many taxa to reasonably plot.

These bar plots are NOT statistical and should be used only as a visual
description of the data (no conclusions should be drawn on specific taxa).
However, it is one of the most common results included in any microbiome paper.

# Part 1: Load packages and data

```{r}
library(tidyverse)
library(phyloseq)
library(microbiome)

ps = readRDS("path/to/save/location/taxonomy_phyloseq.rds")
```

# Part 2: Group taxa and transform counts

We will create a family-level bar plot, where any species that are from the same
family will be grouped.

We will then convert counts to relative abundance.

```{r}
# Group to family level
family = ps %>% tax_glom('Family') %>% 
  # Compositional = relative abundance
  microbiome::transform('compositional')

# All sample counts now sum to 1 (100%).
table(sample_sums(family))
```

# Part 3: Group means and combine low abundances

Next we will calculate the mean abundance of each taxon for each group of
interest (PD and Control). It is ideal to plot one bar per sample, but this is
not feasible for large cohorts.

We will then combine the means for any taxa where mean \< 1%.

```{r}
# 'Melt' the family phyloseq object (create one big data table)
familymelt = family %>% psmelt()

# Mean abundances
familymeans = familymelt %>% 
  group_by(Case_status,Family) %>% # Perform following on every Case_status & Family combo in parallel
  summarize(Abundance = mean(Abundance)) %>% 
  ungroup() # Stop performing in parallel

# Combine <1% (<0.01)
familymeans = familymeans %>% 
  mutate(Family = ifelse(Abundance < 0.01, 'Other (<1%)', Family)) %>% 
  group_by(Case_status,Family) %>% 
  # Combine abundances from all the rare taxa
  summarize(Abundance = sum(Abundance)) %>% 
  ungroup()

# Check that the total for each group still adds up to 1 (100%)
sum(familymeans$Abundance[familymeans$Case_status=='PD'])
sum(familymeans$Abundance[familymeans$Case_status=='Control'])
```

# Part 4: Plot

Results are plotted as a stacked bar plot.

```{r}
familymeans %>% 
  ggplot(aes(Case_status,Abundance,fill=Family)) +
  geom_col(position='stack') +
  theme_classic(base_size=18)
```
