---
title: "Differential Abundance Tutorial"
author: "Avril Metcalfe-Roach"
date: "`r Sys.Date()`"
output: html_document
---

# Background

Differential abundance tests whether differences in the abundances of individual taxa (or pathways/functions) between groups are significant.

Most tools use raw count (or pseudocount) data as input, then apply their own normalization techniques, and then calculate the statistics using some unique statistical approach.

While many tools exist, this tutorial will use ANCOM-BC. Of the high-performing tools, it has middle-of-the-road sensitivity and specificity. it is often recommended to combine the results of multiple tools in order to reduce artefactual bias.

Other recommended tools are MaAsLin2, ALDEx2, and LinDA.

# Part 1: Load libraries and data

```{r}
library(tidyverse)
library(phyloseq)
library(ANCOMBC)
library(microbiome)

ps = readRDS("path/to/save/location/taxonomy_phyloseq.rds")
```

# Part 2: Run tool

ANCOM-BC, like many differential abundance tools, is designed to perform its entire workflow within a single command.

```{r Run_tool}
ancom = ancombc2(ps,
                 tax_level = 'Family',
                 # Fixed-effect (normal) variables. a+b format.
                 fix_formula = 'Case_status',
                 # Benjamini-Hochberg adjustments are by far the most common
                 p_adj_method = 'BH',
                 # Present in at least 30% of samples
                 # Note: no native abundance filter in ANCOM-BC, filter manually if desired
                 prv_cut = 0.3,
                 # FYI: min sampling depth required (default=0)
                 lib_cut = 0,
                 # If your variable of interest is categorical with 3+ groups,
                 # you can set group='variable' and struc_zero=T.
                 # Essentially, this will automatically mark a taxon as significant
                 # if there is zero abundance in at least one (but not all) of the groups.
                 group = NULL,
                 struc_zero = F)
```

ANCOM-BC outputs many statistics and summaries in addition to the significance tests. This section will extract the relevant statistics and form a nicely formatted table.

For this example, it would be easy enough to extract the P adjusted column and save it; however, this wouldn't be very efficient with multivariable results due to the default formatting of ancombc. Instead, the below will work for any number of variables.

Significance is automatically set at Padj<0.05. This is marked by diff=T. Furthermore, if it passes sensitivity analysis (still significant after adding a small pseudocount to zeros, good for reducing artefactually significant rare taxa), this will be marked by passed=T. 

```{r Curate_stats}
# Inspect raw output
View(ancom$res)

# Format into table
stats = ancom$res %>% 
  # Take all columns except taxon and turn into long form data
  pivot_longer(cols = -taxon, 
               names_to = 'stat',
               values_to = 'value') %>% 
  # The stat value denotes the type of statistic and the variable that value is referring to.
  # Therefore, separate stat into two columns.
  separate(col=stat, into = c('stat','variable'),sep='_',remove=T,extra = 'merge') %>% 
  # the statistic passed_ss is the only statistic with a '_'. This was used to separate the
  # names above, and so 'ss_' must be removed from the variable column where stat=='passed'.
  mutate(variable = ifelse(stat=='passed',str_sub(variable,start=4),variable)) %>% 
  # Remove intercept values, not typically useful
  filter(variable != '(Intercept)') %>% 
  # Finally, pivot wider so that each stat value has its own column.
  pivot_wider(names_from = 'stat', values_from = 'value') %>% 
  # Sort by adjusted p values (smallest first)
  arrange(q)

# Inspect the output:
View(stats)

# There are 4 significant family-level microbes. All are higher in PD (lfc>0).
```

# Part 3: Plot results

We will extract the significant families and plot them using both relative abundance and CLR transformations. Both are commonly used.

```{r}
# Extract significant families
sig = stats %>% filter(diff==T, passed==T)

# Generate dataset for plotting
relab = ps %>% 
  tax_glom('Family') %>% 
  # Perform rel ab transformation BEFORE subsetting taxa
  microbiome::transform('compositional') %>% 
  # Subset to only include significant taxa
  subset_taxa(Family %in% sig$taxon) %>% 
  # 'Melt' to transform into data table
  psmelt()

# Plot taxa
relab %>% 
  ggplot(aes(Case_status,Abundance,fill=Case_status)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(height=0, width=0.2) +
  theme_classic(base_size=18) +
  facet_wrap('Family', scales = 'free')

# Let's try it again with CLR transformation

# Generate dataset for plotting
clr = ps %>% 
  tax_glom('Family') %>% 
  # Perform clr transformation BEFORE subsetting taxa
  microbiome::transform('clr') %>% 
  # Subset to only include significant taxa
  subset_taxa(Family %in% sig$taxon) %>% 
  # 'Melt' to transform into data table
  psmelt()

# Plot taxa - better resolution
clr %>% 
  ggplot(aes(Case_status,Abundance,fill=Case_status)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(height=0, width=0.2) +
  theme_classic(base_size=18) +
  facet_wrap('Family', scales = 'free')
```

